Staged Abstract Interpreters

### 2019-09-08 paper/40_absinterp.tex
typo
revision
typos

It is important that the abstract stores now map addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
runtime values.  This can be justified by the approximation nature of abstract
interpretation and nondeterminism in the abstract interpreter.  First, when
analyzing a conditional expression, we may not have enough information to
decide which branch will be taken, thus a sound treatment is to explore the both
branches. Also, at some later point, we need to compute the join of two path.
Another nondeterminism comes from function applications, for instance, @f(a)@,
the possible callees ($\lambda$ terms) of @f@ is computed approximately.  As a
result, we may retrieve a set of closures when dereferencing the address of @f@
and we need to apply all of them.

@@

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is that now the first argument @funs@ is a set of closures (remember
that type @Value@ is an alias of @Set[AbsValue]@), so we can not directly
destruct it to syntactic @Lam@ by pattern matching. Instead, we use an
auxiliary function @lift_nd@ that takes a set and lifts the elements in the set
into the monad stack. Then we can straightforwardly implement the @ap_clo@ in
monadic style, where the closures come from the @SetT@ and thus the
nondeterminism can be naturally handled. The light gray code shows what is
added from the concrete
@ap_clo@.
\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(funs: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](funs)}!
    α <- alloc(x)
    _ <- set_store(α → funs)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination.
The caching algorithm is also called \textit{co-inductive} caching or
\textit{truncated depth-first evaluation} \cite{Rosendahl:AbsIntPL}. It has
been used in other abstract interpreter or fixed point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use a @in@ cache and a @out@ cache during
the depth-first evaluation. The @in@ cache stores the result from the last
iteration, and @out@ cache is used for updating in the current iteration. In the next
iteration, the last @out@ cache will be served as @in@ cache, and an empty cache
is plugged-in to the @out@ slot. Once the @out@ cache does not contain anything
new than the @in@ cache, the fixed-point is reached.

@@

The iteration terminates when the resulting @out@ cache is equivalent to the
input @in@ cache, which indicates that there is no more fact can be discovered.
Therefore, the iteration should end, and we have reached the least fixed point.

\begin{figure}[h!]
  \centering
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache; val cfg = (e, ρ, σ)
    rt <- if (out.contains(cfg)) for { // ask if out already contains the desired result
            (v, σ) <- lift_nd[(Value, Store)](out(cfg));  _ <- put_store(σ)
          } yield v
          else for {
            _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
            v <- ev(fix(ev))(e);  σ <- get_store;  _ <- update_out_cache(cfg, (v, σ))
          } yield v
  } yield rt
\end{lstlisting}
\vspace{-1em}
\caption{The unstaged co-inductive caching algorithm.}
\label{fig:coind_cache}
\vspace{-1em}
\end{figure}

\begin{lstlisting}
  def run(e: Expr): Result = fix(eval)(e)(ρ0)(σ0)(cache0)(cache0).run
\end{lstlisting}

Finally, we override the top-level @run@ method by running the monadic value
with an empty abstract environment $\rho_0$, initial abstract store $\sigma_0$,
and initial caches @cache@$_0$.

>>>>

It is important that the abstract stores now map addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
runtime values.  This can be justified by the approximation nature
and nondeterminism in the abstract interpreter.  First, when
analyzing a conditional expression, we may not have enough information to
decide which branch will be taken, thus a sound treatment is to explore the both
branches. Also, at some later point, we need to compute the join of two path.
Another nondeterminism comes from function applications, for instance, @f(a)@,
the possible callees ($\lambda$ terms) of @f@ is computed approximately.  As a
result, we may retrieve a set of closures when dereferencing the address of @f@
and we need to apply all of them.

@@

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is that now the first argument @funs@ is a set of closures (remember
that type @Value@ is an alias of @Set[AbsValue]@), so we can not directly
destruct it to a syntactic @Lam@ by pattern matching. Instead, we use an
auxiliary function @lift_nd@ that takes a set and lifts the elements in the set
into the monad stack. Then we can straightforwardly implement the @ap_clo@ in
monadic style, where the closures come from the @SetT@ and thus the
nondeterminism can be naturally handled. The light gray code shows what is
added from the concrete
@ap_clo@.
\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(funs: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](funs)}!
    α <- alloc(x)
    _ <- set_store(α → funs)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

@@

As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination.
The caching algorithm is also called \textit{co-inductive} caching or
\textit{truncated depth-first evaluation} \cite{Rosendahl:AbsIntPL}. It has
been used in other abstract interpreters or fixed-point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use a @in@ cache and a @out@ cache during
the depth-first evaluation. The @in@ cache stores the result from the last
iteration, and @out@ cache is used for updating in the current iteration. In the next
iteration, the last @out@ cache will be served as @in@ cache, and an empty cache
is plugged-in to the @out@ slot. Once the @out@ cache does not contain anything
new than the @in@ cache, the fixed-point is reached.

@@

The iteration terminates when the resulting @out@ cache is equivalent to the
input @in@ cache, which indicates that there is no more fact can be discovered.
Therefore, the iteration should end, and we have reached the least fixed-point.
\begin{lstlisting}
  def run(e: Expr): Result = fix(eval)(e)(ρ0)(σ0)(cache0)(cache0).run
\end{lstlisting}

Finally, we override the top-level @run@ method by running the monadic value
with an empty abstract environment $\rho_0$, initial abstract store $\sigma_0$,
and initial caches @cache@$_0$.
\begin{figure}[t]
  \centering
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache; val cfg = (e, ρ, σ)
    rt <- if (out.contains(cfg)) for { // ask if out already contains the desired result
            (v, σ) <- lift_nd[(Value, Store)](out(cfg));  _ <- put_store(σ)
          } yield v
          else for {
            _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
            v <- ev(fix(ev))(e);  σ <- get_store;  _ <- update_out_cache(cfg, (v, σ))
          } yield v
  } yield rt
\end{lstlisting}
\vspace{-1em}
\caption{The unstaged co-inductive caching algorithm.}
\label{fig:coind_cache}
\vspace{-1em}
\end{figure}
